Lesson 3
========================================================
pf <- read.csv('pseudo_facebook.tsv', sep = '\t')

names(pf)

***

### What to Do First?
Notes:

***

### Pseudo-Facebook User Data
Notes:

```{r Pseudo-Facebook User Data}
pf <- read.csv('pseudo_facebook.tsv', sep = '\t')

names(pf)
```

***

### Histogram of Users' Birthdays
Notes:

```{r Histogram of Users\' Birthdays}
library(ggplot2)
ggplot(aes(x = dob_day), data = pf) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous(breaks = 1:31)

```


***

#### What are some things that you notice about this histogram?
Response: Day 1 was really high, the 31st was a bit low but all other days were evenly distributed.

***

### Moira's Investigation
Notes:
She was checking to see how big of an audience people thought they had for a FB post.
***

### Estimating Your Audience Size
Notes:
I no longer post to FB.
***

#### Think about a time when you posted a specific message or shared a photo on Facebook. What was it?
Response:
I don't remember
#### How many of your friends do you think saw that post?
Response:
I don't know
#### Think about what percent of your friends on Facebook see any posts or comments that you make in a month. What percent do you think that is?
Response:
No idea.
***

### Perceived Audience Size
Notes:

***
### Faceting
Notes: facet_grid(vert ~ horiz) data presentation changes on how the var is set. Use for 2 or more var.

```{r Faceting}
ggplot(data = pf, aes(x= dob_day)) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous(breaks = 1:31) +
  facet_wrap(~dob_month)
```

#### Let’s take another look at our plot. What stands out to you here?
Response:
Most entries are for Jan. 1.  This could be people accepting defaults.
***

### Be Skeptical - Outliers and Anomalies
Notes:

***

### Moira's Outlier
Notes:
#### Which case do you think applies to Moira’s outlier?
Response:

***

### Friend Count
Notes:

#### What code would you enter to create a histogram of friend counts?

```{r Friend Count}
library(ggplot2)
ggplot( data = pf, aes(x = friend_count)) +
  geom_histogram() 
  
```

#### How is this plot similar to Moira's first plot?
Response:

***

### Limiting the Axes
Notes:

```{r Limiting the Axes}

```

### Exploring with Bin Width
Notes:

***

### Adjusting the Bin Width
Notes:
breaks(start, end, interval)
### Faceting Friend Count
```{r Faceting Friend Count}
# What code would you add to create a facet the histogram by gender?
# Add it to the code below.
qplot(x = friend_count, data = pf, binwidth = 10) +
  scale_x_continuous(limits = c(0, 1000),
                     breaks = seq(0, 1000, 50))
```
### ggplot equivalent
```{r ggplot equivalent}
ggplot(aes(x = friend_count), data = pf) +
  geom_histogram(binwidth = 25) +
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50))
```
***
### ggplot with gender var
```{r ggplot for gender}
ggplot(aes(x = friend_count), data = pf) +
  geom_histogram() +
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50)) +
  facet_wrap(~gender)
  
```
***
### Omitting NA Values
Notes:
R takes data that does not meet criteria and sets it to NA.
```{r Omitting NA Values}
ggplot(aes(x = friend_count), data = subset(pf, !is.na(gender))) +
  geom_histogram() +
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50)) +
  facet_wrap(~gender)
```

***

### Statistics 'by' Gender
Notes:

```{r Statistics \'by\' Gender}
table(pf$gender)
by(pf$friend_count, pf$gender, summary)
```

#### Who on average has more friends: men or women?
Response:
women
#### What's the difference between the median friend count for women and men?
Response:
22
#### Why would the median be a better measure than the mean?
Response:
median is the actual middle number while mean is adding it all together and averaging.  a few people can skew the mean.
***

### Tenure
Notes:

```{r Tenure}
ggplot(aes(x = tenure), data = pf) +
  geom_histogram(binwidth = 30, color = 'black', fill = '#099dd9')
```

***

#### How would you create a histogram of tenure by year?

```{r Tenure Histogram by Year}
ggplot(aes(x = tenure/365), data = pf) +
  geom_histogram(binwidth = .25, color = 'black', fill = '#f79420') +
  scale_x_continuous(limits = c(0, 7), breaks = seq(1, 7, 1))

```

***

### Labeling Plots
Notes:

```{r Labeling Plots}
ggplot(aes(x = tenure/365), data = pf) +
  geom_histogram(binwidth = .25, color = 'black', fill = '#f79420') +
  scale_x_continuous(limits = c(0, 7), breaks = seq(1, 7, 1)) +
  xlab('Number of years using Facebook') +
  ylab('Number of users in sample')
```

***

### User Ages
Notes:

```{r User Ages}
ggplot(aes(x = age), data = pf) + 
  geom_histogram(binwidth = 1, color = 'black', fill = '#4b0082') +
  scale_x_continuous(limits = c(0, 110), breaks = seq(1, 100, 9))
```

#### What do you notice?
Response:

***

### The Spread of Memes
Notes:

***

### Lada's Money Bag Meme
Notes:

***

### Transforming Data
Notes:

***

### Add a Scaling Layer
Notes:

```{r Add a Scaling Layer}
library(gridExtra)
ap <- ggplot(aes(x = friend_count +1), data = pf) + 
  geom_histogram(color = 'black', fill = '#4b0082')
apl <- ap + scale_x_log10()
aps <- ap + scale_x_sqrt()

grid.arrange(ap, apl, aps, ncol = 1)
```

***


### Frequency Polygons
Notes: 
```{r Frequency Polygons}
ggplot(aes(x = friend_count, y = ..count../sum(..count..)), data = subset(pf, !is.na(gender))) +
  geom_freqpoly(aes(color = gender), binwidth = 10) +
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50)) +
  xlab('Friend_Count') +
  ylab('Proportion of users with that friend count')
  

```
***
```{r Frequency Polygons2}
ggplot(aes(x = www_likes), data = subset(pf, !is.na(gender))) +
         geom_freqpoly(aes(color = gender)) +
         scale_x_log10()
  

```
***

### Likes on the Web
Notes:

```{r Likes on the Web}
by(pf$www_likes, pf$gender, sum)
```


***

### Box Plots
Notes:

```{r Box Plots}
qplot(x = gender, y = friend_count, data = subset(pf, !is.na(gender)), geom = 'boxplot') +
  coord_cartesian(ylim = c(0, 1000))

```

#### Adjust the code to focus on users who have friend counts between 0 and 1000.

```{r}

```

***

### Box Plots, Quartiles, and Friendships
Notes:

```{r Box Plots, Quartiles, and Friendships}

```

#### On average, who initiated more friendships in our sample: men or women?
Response:
#### Write about some ways that you can verify your answer.
Response:because the data shows that most friend requests are below 500 we set coord cart limit there.
```{r Friend Requests by Gender}
qplot(x = gender, y = friendships_initiated, data = subset(pf, !is.na(gender)), geom = 'boxplot') + 
  coord_cartesian(ylim = c(0, 500))
```

Response:

***

### Getting Logical
Notes:

```{r Getting Logical}
summary(pf$mobile_likes)

summary(pf$mobile_likes > 0)

mobile_check_in <- NA

pf$mobile_check_in <- ifelse(pf$mobile_likes > 0, 1, 0)
pf$mobile_check_in <- factor(pf$mobile_check_in)
summary(pf$mobile_check_in)

sum(pf$mobile_check_in ==1) / length(pf$mobile_check_in)
```

Response:

***

### Analyzing One Variable
Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!