---
output:
  html_document: default
  pdf_document: default
---
Lesson 4
========================================================

***

### Scatterplots and Perceived Audience Size
Notes:

***

### Scatterplots
Notes:
qplot(age, friend_count, data = pf) can be used because qplot uses x, y format.
ggplot(aes(x = age, y = friend_count), data = pf) +
  geom_point() 

```{r Scatterplots}
library(ggplot2)
pf <- read.csv('pseudo_facebook.tsv', sep = '\t')

qplot(x = age, y = friend_count, data = pf)

```

***

#### What are some things that you notice right away?
Response:
ages extends beyond 90 and users lower than 30 have many more friends.
***

### ggplot Syntax
Notes:

```{r ggplot Syntax}
ggplot(aes(x = age, y = friend_count), data = pf) +
  geom_point() + xlim(30, 90)
```

***

### Overplotting
Notes:
age is a continuous value.  It is expressed as an int in the plot and as such makes the columns line up neatly which is not a natural state.  By adding jitter we introduce noise to the data giving a more realistic reflection of the data dispersion.
```{r Overplotting}
ggplot(aes(x = age, y = friend_count), data = pf) +
  geom_jitter(alpha = 1/20) + 
  xlim(30, 90) 
```

#### What do you notice in the plot?
Response:
The information scatter feels more realistic and a truer reflection of how the data would be represented.  It is also clearer that the age value of 69 is inaccurate.  Younger users do not seem to be as high as they were before.
***

### Coord_trans()
Notes:

```{r Coord_trans()}
ggplot(aes(x = age, y = friend_count), data = pf) +
  geom_jitter(alpha = 1/20) + 
  xlim(30, 90)
```

#### Look up the documentation for coord_trans() and add a layer to the plot that transforms friend_count using the square root function. Create your plot!

```{r}
ggplot(aes(x = age, y = friend_count), data = pf) +
  geom_point(alpha = 1/20) + 
  xlim(30, 90) +
  coord_trans(y = 'sqrt')
```

#### What do you notice?
By adding the sqrt transform it zooms in on the datapoints and we can see that actual count concentration is well below 1000.
***

### Alpha and Jitter
Notes:

```{r Alpha and Jitter}
ggplot(aes(x = age, y = friendships_initiated), data = pf) + 
  geom_jitter(alpha = 1/10, position = position_jitter(h = 0)) +
  coord_trans(y = 'sqrt')
```

***

### Overplotting and Domain Knowledge
Notes:

***

### Conditional Means
Notes:


```{r Conditional Means}
library(dplyr)
age_groups <- group_by(pf, age)

pf.fc_by_age <- summarise(age_groups,
                          friend_count_mean = mean(friend_count),
                          friend_count_median = median(friend_count),
                          n = n())
pf.fc_by_age <- arrange(pf.fc_by_age, age)
head(pf.fc_by_age)
```
```{r Conditional Means alt}
pf %>%
  group_by(age) %>%
  summarise(friend_count_mean = mean(friend_count),
            friend_count_median = median(friend_count),
            n = n()) %>%
  arrange(age)
head(pf.fc_by_age, 20)

```
Create your plot!

```{r Conditional Means Plot}
ggplot(aes(age, friend_count_mean),data = pf.fc_by_age) +
  geom_line()

```

***

### Overlaying Summaries with Raw Data
Notes:
fun.y take a function and applies it 
```{r Overlaying Summaries with Raw Data}
ggplot(aes(x = age, y = friend_count), data = pf) +
  coord_cartesian(xlim = c(13,90)) +
  geom_point(alpha = 0.05,
             position = position_jitter(h = 0),
             color = 'orange') +
  coord_trans(y = 'sqrt') + 
  geom_line(stat = 'summary', fun.y = mean) +
  geom_line(stat = 'summary', fun.y = quantile, fun.args = list(probs = .1),
            linetype = 2, color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile, fun.args = list(probs = .5),
            color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile, fun.args = list(probs = .9),
            linetype = 2, color = 'blue')

```

#### What are some of your observations of the plot?
Response:

***

### Moira: Histogram Summary and Scatterplot
See the Instructor Notes of this video to download Moira's paper on perceived audience size and to see the final plot.

Notes:

***

### Correlation
Notes:

```{r Correlation}
cor.test(pf$age, pf$friend_count, method = 'pearson')
```
```{r Correlation alt}
with(pf, cor.test(age, friend_count, method = 'pearson'))
```
Look up the documentation for the cor.test function.

What's the correlation between age and friend count? Round to three decimal places.
Response: -0.02740737

***

### Correlation on Subsets
Notes:

```{r Correlation on Subsets}
with(subset(pf, age <= 70), cor.test(age, friend_count,
                                     method = 'pearson'))
```

***

### Correlation Methods
Notes:

***

## Create Scatterplots
Notes:

```{r}
ggplot(aes(www_likes_received, y = likes_received), data = pf) +
  geom_point()
```

***

### Strong Correlations
Notes:

```{r Strong Correlations}
ggplot(aes(www_likes_received, y = likes_received), data = pf) +
  geom_point() +
  xlim(0, quantile(pf$www_likes_received, 0.95)) +
  ylim(0, quantile(pf$likes_received, 0.95)) +
  geom_smooth(method = 'lm', color = 'purple')
```

What's the correlation betwen the two variables? Include the top 5% of values for the variable in the calculation and round to 3 decimal places.

```{r Correlation Calcuation}
with(pf, cor.test(www_likes_received, likes_received, method = 'pearson'))
```

Response:

***

### Moira on Correlation
Notes:

***

### More Caution with Correlation
Notes:

```{r More Caution With Correlation}
library(alr3)
data("Mitchell")
?Mitchell
str(Mitchell)
ggplot(aes(x = Month, y = Temp), data = Mitchell) +
  geom_point()
```

Create your plot!

```{r Temp vs Month}
range(Mitchell$Month)
cor.test(Mitchell$Month, Mitchell$Temp, method = 'pearson')
ggplot(aes(x = Month, y = Temp), data = Mitchell) +
  geom_point()
  
```

***

### Noisy Scatterplots
a. Take a guess for the correlation coefficient for the scatterplot.

b. What is the actual correlation of the two variables?
(Round to the thousandths place)

```{r Noisy Scatterplots}

```

***

### Making Sense of Data
Notes:

```{r Making Sense of Data}
range(Mitchell$Month)
cor.test(Mitchell$Month, Mitchell$Temp, method = 'pearson')
ggplot(aes(x = Month, y = Temp), data = Mitchell) +
  geom_point() +
  scale_x_continuous(breaks = seq(0, 203, 12))
```

***

### A New Perspective

What do you notice?
Response:

Watch the solution video and check out the Instructor Notes!
Notes:

***

### Understanding Noise: Age to Age Months
Notes:

```{r Understanding Noise: Age to Age Months}
pf$age_with_months <- pf$age + (12 - pf$dob_month) / 12
```

***

### Age with Months Means

```{r Age with Months Means}
pf.fc_by_age_months <- pf %>%
  group_by(age_with_months) %>%
  summarise(friend_count_mean = mean(friend_count),
            friend_count_median = median(friend_count),
            n = n())%>%
  arrange(age_with_months)
```

Programming Assignment
```{r Programming Assignment}
library(dplyr)
pf.age_with_months <- pf %>%
  group_by(age_with_months) %>%
  summarise(friend_count_mean = mean(friend_count),
            friend_count_median = median(friend_count),
            n = n()) %>%
  arrange(age_with_months)
```

***

### Noise in Conditional Means

```{r Noise in Conditional Means}
ggplot(aes(x = age_with_months, y = friend_count_mean),
       data = subset(pf.age_with_months, age_with_months < 71)) +
  geom_line()
```

***

### Smoothing Conditional Means
Notes:
Bias variant tradeoff.
```{r Smoothing Conditional Means}
p1 <- ggplot(aes(x = age, y = friend_count_mean),
       data = pf.fc_by_age) +
  geom_line()+
  geom_smooth()

p2 <- ggplot(aes(x = age_with_months, y = friend_count_mean),
       data = subset(pf.fc_by_age_months, age_with_months < 71)) +
        geom_line() +
        geom_smooth()

p3 <- ggplot(aes(x = round(age / 5) * 5, y = friend_count),
             data = subset(pf, age < 71)) +
  geom_line(stat = 'summary', fun.y = mean)
library(gridExtra)
grid.arrange(p1, p2, p3, ncol =  1)
```

***

### Which Plot to Choose?
Notes:
You don't have to choose.  We will often create multiple plots and summaries of the data. They can reveal different things about the data.  When publishing you may want to narrow the scope to the plots that best communicate the findings.
***

### Analyzing Two Variables
Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

